---
title: "Exercise 1.1"
output: 
  ioslides_presentation:
    widescreen: true
    transition: faster
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment="")
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(shinystan)
library("ggplot2")

set.seed(1234)
```

```{r setup2, include=F, cache=T}
srrs2 = read.csv("srrs2.dat", strip.white=TRUE)
srrs.mn = srrs2[srrs2$state == "MN",]
srrs.mn$fips = srrs.mn$stfips * 1000 + srrs.mn$cntyfips
cty = read.csv("cty.dat", strip.white=TRUE)
cty.mn = cty[cty$st=="MN",]
cty.mn$fips = 1000 * cty.mn$stfips + cty.mn$ctfips
srrs.mn = merge(srrs.mn, cty.mn[c("fips", "Uppm")], by = 'fips')
srrs.mn = srrs.mn[!duplicated(srrs.mn[,'idnum']),]
u = log(srrs.mn$Uppm)
n = nrow(srrs.mn)
head(srrs.mn)
srrs.mn$county = as.character(srrs.mn$county)
mn.counties = unique(srrs.mn$county)
county.lookup = as.vector(1:length(mn.counties), mode='list')
names(county.lookup) = mn.counties
county = srrs.mn$county.code = unlist(county.lookup[srrs.mn$county], use.names=F)
radon = srrs.mn$activity
srrs.mn$log.radon = log.radon = log(radon + 0.1)
floor.measure = srrs.mn$floor
df = data.frame(log.radon, floor.measure, county)
write.csv(df, file="dist/data.csv", row.names = F)
```

## Radon contamination
- radioactive gas that enters homes through the ground
- carcinogenic
- vary greatly household to household

## EPA Study
- 80,000 houses
> - Two predictors:
>    - measument on lowest floor?
>    - county uranium level (positive corr with radon levels)
> - We'll focus on Minnesota, which has 919 households in 85 counties

## Distribution of radon levels in MN (log scale)
```{r}
ggplot(df, aes(log.radon)) + geom_histogram(bins = 25)
```

## Conventional approaches
* Complete pooling
    - Treat all counties the same and estimate a single radon level,
    $y_i \sim \mathcal{N}(\alpha + \beta x_i, \sigma)$
* No pooling
    - Model radon in each county independently,
    $y_i \sim \mathcal{N}(\alpha_{j[i]} + \beta x_i, \sigma)$, where $j = 1, ..., 85$
    
As usual, we can think of $\sigma$ as capturing all of the variance we aren't modeling.

## Pooled model
```{r}
writeLines(readLines("pooled.stan"))
```

## Pooled fit
```{r, cache=T, include=F}
pooled.model = stan_model("pooled.stan")
pooled.fit = sampling(pooled.model, list(y=df$log.radon,
                           x = df$floor.measure,
                           N=nrow(df)))
```
```{r}
print(pooled.fit, probs = c(0.025, 0.5, 0.975))
```
## Pooled fit
```{r pooled.fit}
pooled.sample = extract(pooled.fit)
b = mean(pooled.sample$beta[,1])
m = mean(pooled.sample$beta[,2])

ggplot(df, aes(floor.measure, log.radon)) + geom_count() +
  geom_abline(intercept = b, slope = m, linetype=2)
```

## Let's try writing an unpooled model
- We want a separate intercept for each county, but we can share beta and sigma.

## Unpooled model
```{r}
writeLines(readLines("unpooled.stan"))
```

## Unpooled fit
```{r unpooled.fit1, cache=T, include=F}
unpooled.model = stan_model("unpooled.stan")
unpooled.fit = sampling(unpooled.model, list(y=df$log.radon,
                           x = df$floor.measure,
                           county = df$county,
                           N=nrow(df)))
```
```{r}
print(unpooled.fit, probs = c(0.025, 0.5, 0.975))
```

## Unpooled fit
```{r unpooled.fit}
unpooled.samples = extract(unpooled.fit)
unpooled.estimates = apply(unpooled.samples$a, 2, mean)
unpooled.se = apply(unpooled.samples$a, 2, sd)
unpooled.df = data.frame(unpooled.estimates, 1:85, 
                         unpooled.estimates + unpooled.se,
                         unpooled.estimates - unpooled.se)
names(unpooled.df) = c("a", "county", "upper", "lower")
unpooled.df = unpooled.df[order(unpooled.df$a),]

ggplot(unpooled.df, aes(x=1:85, y=a)) + geom_pointrange(aes(ymin = lower, ymax = upper))
```
<!--
## Visual comparisons between pooled and unpooled
- TODO ?
-->

## meh
- Pooling is useless for identifying high-radon counties
- We don't trust unpooled estimates especially for counties with little data

## DIY partial pooling
Let's say that each county should get its own intercept $\alpha_j$ and effect of floor measurement $\beta_j$ and code $y_i \sim ~ \mathcal{N}(\alpha_{j[i]} + \beta_{j[i]} x_i, \, \sigma)$

## Varying slopes and intercepts by group
```{r}
writeLines(readLines("partial_max.stan"))
```

```{r, include=F, cache=T}
partial.model = stan_model("partial_max.stan")
partial.fit = sampling(partial.model,
                       list(N = nrow(df), y = df$log.radon,
                            x = df$floor.measure, county = df$county,
                            J = length(unique(df$county)))
                       , control=list(adapt_delta=0.98))
```

## 
