---
title: "Exercise 2.2"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment="", autodep = T)

library(shinystan)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

N = 5
K = 3
X = matrix(rnorm(N*K, 0, 10), ncol=K)
write.csv(X, file = "dist/X.csv", row.names = F)
alpha = rnorm(1, 0, 1)
beta = rnorm(K, 0, 1)
sigma = 4
y = rnorm(N, beta * X + alpha, sigma)
write(y, file="dist/y.txt")
```

## A fun, fairly simple linear regression model
```{r}
writeLines(readLines("model.stan"))
model = stan_model("model.stan")
```

## Problem fit

```{r, cache=T, message=F}
fit = sampling(model, list(y=y, N=N, X=X))
```

## Warnings
- "X divergent transitions; increase adapt_delta"
- "Y transitions exceeded maximum treedepth"
- "Examine the pairs() plot"

## The fit itself
```{r}
fit
```

## pairs() plot
```{r warning=F, error=F}
pairs(fit, pars = c('alpha', 'beta[1]', 'sigma'))
```

## Woops, let's add priors
```{r}
writeLines(readLines("model2.stan"))
model2 = stan_model("model2.stan")
```


## Now with 100% more priors!

```{r, cache=T, message=F}
fit2 = sampling(model2, list(y=y, N=N, X=X))
```
```{r}
fit2
```

## Pairs() plot, plus priors
```{r echo=T}
pairs(fit2, pars = c('alpha', 'beta[1]', 'sigma'))
```

## Raising adapt_delta

```{r, cache=T, message=F, echo=T}
fit2.1 = sampling(model2, list(y=y, N=N, X=X),
               control=list(adapt_delta=0.99))
```

## The more careful fit
- No more divergences!
```{r} 
fit2.1
```

## Punctilious pairs() plot
```{r echo=T}
pairs(fit2.1, pars = c('alpha', 'beta[1]', 'sigma'))
```

## More informative priors
```{r}
writeLines(readLines("model3.stan"))
model3 = stan_model("model3.stan")
```

## Fit
```{r, cache=T, message=F, echo=T}
fit3 = sampling(model3, list(y=y, N=N, X=X))
```

## More informative prior fit
```{r}
fit3
```

## Pairs plot
```{r echo=T}
pairs(fit3, pars = c('alpha', 'beta[1]', 'sigma'))
```

## Home safe?
- No more diagnostic warnings!
- PPCs

## PPC model
```{r}
writeLines(readLines("model3ppc.stan")[-(1:11)])
model3ppc = stan_model("model3ppc.stan")
```

```{r, include=F}
fit3ppc = sampling(model3ppc, list(y=y, N=N, X=X))
```

## Shinystan
```{r, echo=T, eval=F}
library(shinystan)
launch_shinystan(fit3ppc)
```

## So what was that data, anyway?
```{r echo=T}
alpha
beta
sigma
```

## The fit
```{r}
fit3
```

## Non-identified!