---
title: "Exercise 2.2"
output: 
  ioslides_presentation:
    template: tnew.html
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment="", autodep = T, tidy.opts=list(width.cutoff=60), tidy=TRUE)

library(shinystan)
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
set.seed(1234)

N = 5
K = 3
X = matrix(rnorm(N*K, 0, 10), ncol=K)
write.csv(X, file = "dist/X.csv", row.names = F)
alpha = rnorm(1, 0, 1)
beta = rnorm(K, 0, 1)
sigma = 4
y = rnorm(N, beta * X + alpha, sigma)
write(y, file="dist/y.txt")
```

## Let's write another linear regression.
We should know this by now, right? $y \sim \mathcal{N}(\beta X + \alpha, \sigma)$
```
data {
  int N;
  int K;
  matrix[N, K] X;
  real y[N];
}
```

## Full linear regression model
```{r}
writeLines(readLines("model.stan"))
model = stan_model("model.stan")
```

## Problem fit

```{r, cache=T, message=F}
fit = sampling(model, list(y=y, N=N, X=X), seed=1234)
```

## Warnings
- "X divergent transitions; increase adapt_delta"
- "Y transitions exceeded maximum treedepth"
- "Examine the pairs() plot"
- We run multiple chains partially to ensure we find issues

## The fit itself
```{r}
print(fit, probs = c())
```

## pairs() plot
```{r warning=F, error=F}
pairs(fit, pars = c('alpha', 'beta[1]', 'sigma'))
```

## Woops, let's add priors
```{r}
writeLines(readLines("model2.stan"))
model2 = stan_model("model2.stan")
```


## Now with 100% more priors!

```{r, cache=T, message=F}
fit2 = sampling(model2, list(y=y, N=N, X=X), seed=1234)
```
```{r}
fit2
```

## Pairs() plot, plus priors
```{r echo=T}
pairs(fit2, pars = c('alpha', 'beta[1]', 'sigma'))
```

## Raising adapt_delta

```{r, cache=T, message=F, echo=T}
fit2.1 = sampling(model2, list(y=y, N=N, X=X),
               control=list(adapt_delta=0.99), seed=1234)
```

## The more careful fit
- No more divergences!
```{r} 
fit2.1
```

## Punctilious pairs() plot
```{r echo=T}
pairs(fit2.1, pars = c('alpha', 'beta[1]', 'sigma'))
```

## More informative priors
```{r}
writeLines(readLines("model3.stan"))
model3 = stan_model("model3.stan")
```

## Fit
```{r, cache=T, message=F, echo=T}
fit3 = sampling(model3, list(y=y, N=N, X=X), seed=1234)
```

## More informative prior fit
```{r}
fit3
```

## Pairs plot
```{r echo=T}
pairs(fit3, pars = c('alpha', 'beta[1]', 'sigma'))
```

## Home safe?
- PPCs

## PPC model
```{r}
writeLines(readLines("model3ppc.stan")[-(1:11)])
model3ppc = stan_model("model3ppc.stan")
```

```{r, include=F}
fit3ppc = sampling(model3ppc, list(y=y, N=N, X=X), seed=1234)
```

## Shinystan
```{r, echo=T, eval=F}
library(shinystan)
launch_shinystan(fit3ppc)
```

## What parameters were used to generate the data?
```{r echo=T}
alpha
beta
sigma
```

## The fit
```{r}
fit3
```

## What was our data, anyway?
```{r, echo=T}
readLines("dist/y.txt")
```

## Non-identified!
- beta[1:3], alpha, sigma vs. 5 datapoints
- Prior choice can help fit, but make sure it aligns with your understanding of the system
- `adapt_delta` safer to tune

## The more careful fit (less informative priors)
```{r}
writeLines(readLines("model2.stan")[-(1:13)])
fit2.1
```