---
title: "Exercise 1.1"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment="")
height_data = read.csv("heights.csv")
height_data = height_data[height_data$age >= 18, ]
hdata = height_data
library(rstan)
library(shinystan)
library("ggplot2")
set.seed(1234)
```
## Modeling heights in a population
The !Kung San are the most famous foraging population of the 20th century, mostly because they have the most detailed quantitative studies conducted by anthropologists like Nancy Howell in the late 1960s.

I've excluded the children from the data for now, so we're only modeling adult heights.

```{r, echo=T}
str(height_data)
```

## Simple modeling
- How tall is the average !Kung San?
```{r, echo=T}
mean(height_data$height)
```
- How dispersed are the heights?
```{r, echo=T}
sd(height_data$height)
```
- What might we expect a new !Kung San height to look like?
```{r, echo=T}
rnorm(1, mean(height_data$height), sd(height_data$height))
```
## Draws all the way down
- Primary output of a fit object is a set of draws of parameter values
- Draws retain uncertainty, correlation, and other aspects of the posterior that point estimates lose
- Standard toolbox for working with draws
- Summarize only when required

## Bayesian ways to summarize a distribution
```{r, message=F, warning=F, error=F, cache=T}
plot_title <- ggtitle("Height and weight distributions",
                      "with medians and 80% intervals")
mcmc_areas(height_data,
           pars = c("height", "weight"),
           prob = 0.8) + plot_title
```

## Resources
- http://mc-stan.org > Users > Documentation > Stan Manual (pdf)

```{r}
writeLines(readLines("heights1.stan")[1:4])
```
```
parameters { ??? }
model { ??? }
```

## A first draft
```{r}
writeLines(readLines("heights1.stan"))
```
```{r results='hide', cache=T}
model1 = stan_model("heights1.stan")
```
- Priors

## The fit
```{r, results='hide', cache=T}
fit1 = sampling(model1, list(num_people=nrow(hdata),
                             heights=hdata$height))
```
```{r}
print(fit1, prob = c(0.025, .50, 0.975))
```

## Let's see if we can get a better fit
We'll look at better ways of evaluating this later, but for now we'll try to reduce the standard deviation for height by introducing weight as a predictive feature and adding a regression coefficient for it.

```{r}
writeLines(readLines("heights2.stan")[1:5])
```

## Weight regression model
```{r}
writeLines(readLines("heights2.stan")[-(1:5)])
```

## Weight regression fit
```{r, results='hide', cache=T}
model2 = stan_model("heights2.stan")
fit2 = sampling(model2, list(num_people = nrow(hdata), heights = hdata$height,
                             weights = hdata$weight))
```
```{r}
print(fit2, probs = c(0.025, 0.975))
```

## Interpretation
- height_location_avg_weight
- weight_coeff
```{r, echo=T}
sd(height_data$weight) #kg
```

## Unidentified models have correlations
```{r, echo=T}
pairs(fit2, pars = c("height_location_avg_weight", "weight_coeff"))
```